---
title: "Spatial Models"
author: "Morgan Will"
date: "2/22/2022"
output: word_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir='../')
```

## Spatial Modeling Assignment

```{r}
library(vegan)
library(nlme)
data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

1) Examine if there is evidence of spatial dependence in a rare and a common
species in the BCI tree dataset

```{r}
#All Data
BCI_rowsum<-rowSums(BCI>0)
hist(BCI_rowsum)
BCI_col_brks <- hist(BCI_rowsum, plot=F)$breaks
BCI_col_indices <- as.numeric(cut(BCI_rowsum, BCI_col_brks))
BCI_cols <- rev(terrain.colors(length(BCI_col_brks)))
plot(BCI_xy, cex=2, pch=19, col=BCI_cols[BCI_col_indices])
```

```{r}
#Sum columns to determine number of observations of each species
BCI_colsum <- colSums(BCI > 0)
#Evaluate colsum values
quantile(BCI_colsum)
hist(BCI_colsum)
plot(density(BCI_colsum))
#Identify a "rare" and a "common" species
which(BCI_colsum > 10 & BCI_colsum < 14)
which(BCI_colsum > 37 & BCI_colsum < 40)
#Rare Species: Anacardium.excelsum, 13 individuals recorded
#Common Specie: Inga.sapindoides, 38 individuals recorded
```
```{r}
#Evaluate spatial dependence in rare and common species
rare <- BCI[,13]
common <- BCI[,110]
#Rare Plots
hist(rare)
rare_col_brks <- hist(rare, plot=F)$breaks
rare_col_indices <- as.numeric(cut(rare, rare_col_brks))
rare_cols <- rev(terrain.colors(length(rare_col_brks)))
plot(BCI_xy, cex=2, pch=19, col=rare_cols[rare_col_indices])
```
```{r}
#Common Plots
hist(common)
common_col_brks <- hist(common, plot=F)$breaks
common_col_indices <- as.numeric(cut(common, common_col_brks))
common_cols <- rev(terrain.colors(length(common_col_brks)))
plot(BCI_xy, cex=2, pch=19, col=common_cols[common_col_indices])
```
Spatial dependence does not seem to be present in either of the species evaluated (rare and common)

2) Build two generalized linear models to predict the abundance of the species
*Drypetes standleyi* using the abundance of other tree species in the study site.
Specifically examine the following species as predictor variables:

```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")
head(BCI)
```
Note renaming the species ids to something a little easier to work with like
"sp_a", "sp_b" will make model construction a little less cumbersome

```{r}
sp_i<-BCI[,63]  #Drypetes.standleyi
sp_a<-BCI[,51]  #Cordia.lasiocalyx
sp_b<-BCI[,97]  #Hirtella.triandra
sp_c<-BCI[,152] #Picramnia.latifolia
sp_d<-BCI[,171] #Quassia.amara
sp_e<-BCI[,193] #Tabernaemontana.arborea
sp_f<-BCI[,204] #Trattinnickia.aspera
sp_g<-BCI[,220] #Xylopia.macrantha
```

* Model 1: only include a single species as a predictor variable
```{r}
#Visualize abundance of species i across sites
col_brks<-hist(sp_i, plot=F)$breaks
col_indices<-as.numeric(cut(sp_i, col_brks))
cols<-rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

#Visualize abundance of species a across sites
col_brks<-hist(sp_a, plot=F)$breaks
col_indices<-as.numeric(cut(sp_a, col_brks))
cols<-rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

#Model comparing abundance of species i to abundance of species a
SingleSpModel<-glm(sp_i~sp_a, family="gaussian")
SingleSpModel
#abundance of species a at a given site does not correlate with abundance of species i at a site, so species a is not a useful predictor variable for modeling abundance of species i

#Is there another species in the group that is a better predictor of species i abundance?
glm(sp_i~sp_a+sp_b+sp_c+sp_d+sp_e+sp_f+sp_g, family="gaussian")
#none of the species in the group returned values indicating significant correlation with species i
```

* Model 2: include all of the species as predictor variables
```{r}
#Visualize abundance of species i across sites
col_brks<-hist(sp_i, plot=F)$breaks
col_indices<-as.numeric(cut(sp_i, col_brks))
cols<-rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

#Visualize abundance of group of species (a through g)
pred_sp<-data.frame(sp_a, sp_b, sp_c, sp_d, sp_e, sp_f, sp_g)
pred_sp_rs<-rowSums(pred_sp>0)
col_brks<-hist(pred_sp_rs, plot=F)$breaks
col_indices<-as.numeric(cut(pred_sp_rs, col_brks))
cols<-rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])

#Model comparing abundance of species i to abundance of group of species (a through g)
MultiSpModel<-glm(sp_i~pred_sp_rs, family="gaussian")
MultiSpModel
#this model also shows no significant correlation between species i abundance and species group abundance
```

With both models examine the spatial dependence of the residuals using the 
function `Variogram`. Model the spatial dependence in the residuals using one
of the error structures available. 
```{r}
?Variogram
#Single Species Model: Initial Variogram
sr_dat <- data.frame(BCI, BCI_xy)
sr_lm <- gls(Drypetes.standleyi~Cordia.lasiocalyx, data=sr_dat)
plot(Variogram(sr_lm, form=~x+y))

#Single Species Model: Exponential Variogram
sr_exp <- update(sr_lm, corr=corExp(form=~x+y))
xy_dist<- dist(BCI_xy)
max_dist <- max(xy_dist)/2
plot(Variogram(sr_exp, maxDist = max_dist))

#Single Species Model: Normalized Exponential
plot(Variogram(sr_exp, resType='normalized', maxDist=max_dist))

#Single Species Model: Rational Quadratic
sr_rat_nug <- update(sr_lm, corr=corRatio(form=~x+y, nugget=T))
plot(Variogram(sr_rat_nug, maxDist=max_dist))

#Single Species Model: Normalized Rational Quadratic
plot(Variogram(sr_rat_nug, resType='n', maxDist=max_dist))

#Compare models:
anova(sr_lm, sr_exp, sr_rat_nug)
```
```{r}
#Multiple Species Model: Initial Variogram
sr_dat <- data.frame(BCI, BCI_xy)
sr_lm_2 <- gls(Drypetes.standleyi~Cordia.lasiocalyx+Hirtella.triandra+Picramnia.latifolia+Quassia.amara+Tabernaemontana.arborea+Trattinnickia.aspera+Xylopia.macrantha, data=sr_dat)
plot(Variogram(sr_lm_2, form=~x+y))

#Multiple Species Model: Exponential Variogram
sr_exp_2 <- update(sr_lm_2, corr=corExp(form=~x+y))
xy_dist<- dist(BCI_xy)
max_dist <- max(xy_dist)/2
plot(Variogram(sr_exp_2, maxDist = max_dist))

#Multiple Species Model: Normalized Exponential
plot(Variogram(sr_exp_2, resType='normalized', maxDist=max_dist))

#Multiple Species Model: Rational Quadratic
sr_rat_nug_2 <- update(sr_lm_2, corr=corRatio(form=~x+y, nugget=T))
plot(Variogram(sr_rat_nug_2, maxDist=max_dist))

#Multiple Species Model: Normalized Rational Quadratic
plot(Variogram(sr_rat_nug_2, resType='n', maxDist=max_dist))

#Compare models:
anova(sr_lm_2, sr_exp_2, sr_rat_nug_2)
```
* Did including the spatial error term have a large impact on the coefficients
of the model?
The t-values in the models changed a fair amount when spatial error was taken into account. Despite this, the single species model still did not show any significant correlation, while the exponential multi-species model resulted in significant results.

* Did including the spatial error terms significantly improve model fit (use 
function `anova` to carry out model comparison)?
Yes, though the incorporation of spatial error did not make all of the resulting models significant, the models with spatial error considerations (sr_exp and sr_rat_nug) were better fits for the data than those without spatial considerations (sr_lm) in both the single species and multi-species models. This evaluation was made based on the resulting AIC values when evaluated using the anova function (lower AIC indicates better fit).

* Explain why you did or did not observe a difference in the influence of adding the spatial error term between the two models.
A difference was seem because taking the spatial value into consideration reduces some of the autocorrelation found in the model. Though explanatory variables can account for large portions of variability in species distribution, there will always be an underlying factor of range and distribution capability to take into account. In other words, species may simply be grouped together because they did not travel far from the parent, not due to an environmental explanation. This is the variability that is taken into account when adding spatial error to the model. 
